Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-03 paper.qmd
some updates to method section

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
This work will use H3 cells at resolution 3 which corresponds to hexagons of an
average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.

This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provides a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematical background for the evaluation of
collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

>>>>

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.

      <!-- is there more to say and provide context? -->

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
However, this paper augments the approximation of risk exposure. 
Instead of approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

This work focusses on the European region.
The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1].
To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
The reduction of the spatial complexity enables fast processing.
For this work, we employ a H3-cell-resolution of 3. 
This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->

@@

To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
The flight densities are characterised on an hourly basis for all days in these weeks.

<!-- next sentence feels like an orphan ... do we need this here? ------>
This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
